---
title:  'Final Project Worked Exampple'
subtitle: 'INFO 523'
author:
- name: Student -  [Danielle Stea]
  affiliation: [AZ Online]
- name: Instructor -  Cristian Román-Palacios
  affiliation: School of Information, University of Arizona, Tucson, AZ
output: html_document

---
This example goes through the application of principal components analysis to gene expression data from patients with acute myeloid leukemia (AML) and acute lymphoblastic leukemia (ALL). This data was published in 1999 by Gulub et al. and is shared publicly on Kaggle. The dataset can be found at https://www.kaggle.com/datasets/crawford/gene-expression?datasetId=1868

```{r, echo=FALSE, results='hide', message=FALSE}
# loads in the necessary libraries
library(tidyverse)

```
The following tutorial was followed to create the PCA script: https://www.youtube.com/watch?v=0Jp4gsfOLMs

This section serves to bring in and clean up the relevant data.
```{r}
# reads in the training dataset dataset from the downloaded files
train <- read.csv("Data/data_set_ALL_AML_train.csv")
# renames the row names with the gene number
row.names(train) <- train$Gene.Accession.Number
# removes 'call' and Gene description columns from the dataset and any rows with NA values
train <- train %>%
  select(-contains(c('call','Gene'))) %>%
  na.omit() 
#reads in the list of patients and their cancer diagnosis
patients_list <- read.csv("Data/actual.csv")
#Removes the first letter of each column name (the X) and creates a list with those column names
names(train) <- substring(names(train),2)
train <- train %>% 
  select(order(as.numeric(colnames(.))))
p1 <- colnames(train)
#filters out the patients not used in the training dataset
patients_list <- patients_list %>% filter(row_number() %in% p1)
#creates a new column that merges the patient number with the cancer abbreviation
patients_list <- patients_list %>%
  unite(description, c("cancer","patient"))

# renames the columns in the training dataset to display the patient number and the diagnosed cancer type
names <- patients_list$description
colnames(train) <- names
```


This section performs the necessary PCA calculations.
```{r}
#creates a matrix from the train dataframe
train_matrix <- as.matrix(train)
# performs pca on the transposed and scaled train matrix
pca <- prcomp(t(train_matrix), scale = TRUE)
#calculates the squared standard deviation to determine the variation of each principal component
pca_var <- pca$sdev^2
#calculates the percentage of variation of each principal component
var_percent <- 100*pca_var/sum(pca_var)

#creates a barplot with the most significant principal components
barplot(var_percent, xlab = "Principal Component", ylab = "Percent of Variation")
```

```{r}
#makes the PCA data into a dataframe in order to make plots with it
pca_df <- data.frame(Sample = rownames(pca$x),
                    PC1 = pca$x[,1], PC2 = pca$x[,2], PC3 = pca$x[,3])

#Plots PC1 vs. PC2
ggplot(pca_df, aes(x=PC1, y=PC2, label = Sample))+
  geom_text()

#Plots PC1 vs PC3
ggplot(pca_df, aes(x=PC1, y=PC3, label = Sample))+
  geom_text()

```

```{r}
#sorts the loading scores of the PCA by absolute value, gives a list of the ten genes contributing the most to variation in PC1.
top_genes_PC1 <- names(sort(abs(pca$rotation[,1]), decreasing = TRUE)[1:10])
print("10 Genes contributing most to variation in PC1")
pca$rotation[top_genes_PC1,1]

```